---
title: "Sade Feldman Analysis"
author: "Kari Stromhaug"
date: "11/6/2020"
output: html_document
---

```{r setup, include=FALSE}
library(Seurat)
library(tidyverse)
library(splitstackshape)
library(openxlsx)
library(reshape2)
library(pheatmap)
source('~/Documents/source_code/seurat_functions.R')

setwd('~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/Sade-Feldman/')

```

```{r heatmap function, message=F}
MakeHeatmapTable <- function(obj, internal_label, external_label) {
  temp = data.frame(table(obj@meta.data[,c(internal_label, external_label)]))
  temp = dcast(temp, paste0(internal_label, ' ~ ', external_label))

  rownames(temp) <- temp[,internal_label]
  temp[,internal_label]<-NULL

  temp.t = t(temp)
  temp.tp = temp.t
  
  for (i in 1:nrow(temp.tp)) {
    temp.tp[i,] <- temp.tp[i,] / sum(temp.tp[i,]) * 100
  }
  for (i in 1:ncol(temp.tp)) {
    temp.tp[,i] <- temp.tp[,i] / sum(temp.tp[,i]) * 100
  }
  
  return(temp.tp)
}

```


```{r majclust}
AddMajorityCluster <- function(obj, cluster_name) {
  show('ADDING MAJORITY CLUSTER')
  cols = c(cluster_name, 'clonotype.family.3')
  
  complete.table = obj@meta.data[,cols]
  
  unique.clones = unique(complete.table$clonotype.family.3); cat('number of unique clones:', length(unique.clones), '\n')
  
  unique.clones = unique.clones[!is.na(unique.clones) & !grepl('-NA', unique.clones)]; cat('number that are not NA:', length(unique.clones), '\n')

  for (clones in unique.clones) {
    # cat('the clone is:', clones, '\n')
    clones.sub = complete.table %>% subset(clonotype.family.3==clones)
    # cat('dimensions of subset:', dim(clones.sub), '\n')
    cluster.dist = table(clones.sub[,cluster_name])
    num.clones = nrow(clones.sub)

    cluster.dist <- data.frame(cluster.dist)
    rownames(cluster.dist) <- cluster.dist[,1]
    cluster.dist[,1] <- NULL

    maj.clust = rownames(cluster.dist)[apply(cluster.dist,2,which.max)]
    
    mult = cluster.dist %>% subset(Freq == max(cluster.dist$Freq))
    if (nrow(mult) == 2) {
      maj.clust = '2.multiple'
    } else if (nrow(mult) >= 3) {
      maj.clust = '3.or.more'
    }
    # cat('majority cluster:', maj.clust, '\n')

    complete.table[which(complete.table$clonotype.family.3==clones), paste0('maj.', cluster_name)] <- maj.clust
  }
  return(complete.table)
}
```

```{r read dataF}
sfs <- readRDS('s_f_seurat_unprocessed.rds')

## read in protein-coding genes
protein_genes = readRDS('~/Dropbox (Partners HealthCare)/reference_data/gene_lists/protein_coding_genes_pseudogenes_list.rds')

length(intersect(protein_genes, rownames(sfs)))

```


## Seurat processing using their labels

```{r process cd8, message=F}
sfs = readRDS('s_f_seurat_unprocessed.rds')
sfs@meta.data = sfs@meta.data[colnames(sfs),]
Idents(sfs) <- sfs$Cell.type
just.cd8 = subset(sfs, idents='CD8')

just.cd8 = just.cd8[intersect(protein_genes, rownames(just.cd8)), ]

just.cd8 <- DoSeurat(just.cd8, do_norm=F, res=0.45)
just.cd8$clusters <- just.cd8$seurat_clusters
# just.cd8 <- FindClusters(just.cd8, resolution =0.4)
# ```{r cluster res, echo=F, eval=F, message=F}
# just.cd8 <- readRDS('s_f_cd8all_res06_20200901.rds')

```


```{r resolutions, echo=F, eval=F}
just.cd8 <- just.cd8 %>% FindClusters(res=0.35)
just.cd8$clusters_res_0.35 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.37)
just.cd8$clusters_res_0.37 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.40)
just.cd8$clusters_res_0.40 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.42)
just.cd8$clusters_res_0.42 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.45)
just.cd8$clusters_res_0.45 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.50)
just.cd8$clusters_res_0.50 <- just.cd8$seurat_clusters

just.cd8 <- just.cd8 %>% FindClusters(res=0.60)
just.cd8$clusters_res_0.60 <- just.cd8$seurat_clusters

FeaturePlot(just.cd8, features=c('PDCD1', 'ENTPD1', 'TCF7', 'MKI67', 'IL7R'))

DimPlot(just.cd8, group.by='clusters_res_0.30')
DimPlot(just.cd8, group.by='clusters_res_0.35')
DimPlot(just.cd8, group.by='clusters_res_0.37')
DimPlot(just.cd8, group.by='clusters_res_0.40')
DimPlot(just.cd8, group.by='clusters_res_0.42')
DimPlot(just.cd8, group.by='clusters_res_0.45')
DimPlot(just.cd8, group.by='clusters_res_0.50')
DimPlot(just.cd8, group.by='clusters_res_0.60')


just.cd8 <- just.cd8 %>% FindClusters(res=0.80)
just.cd8$clusters_res_0.80 <- just.cd8$seurat_clusters
DimPlot(just.cd8, group.by='clusters_res_0.80')

just.cd8 <- just.cd8 %>% FindClusters(res=1)
just.cd8$clusters_res_1.0 <- just.cd8$seurat_clusters
DimPlot(just.cd8, group.by='clusters_res_1.0')

just.cd8$clusters <- just.cd8$clusters_res_0.45
```

## Moshe's cluster labels
```{r }
# DimPlot(just.cd8, label=T)
DimPlot(just.cd8, group.by='CD8.cluster...6.clusters')
DimPlot(just.cd8, group.by='CD8.cluster...2.clusters')
# FeaturePlot(just.cd8, features=c('CD3D', 'PTPRC', 'CD4', 'CD8A'))

# saveRDS(just.cd8, 's_f_prelabeledcd8_processed.rds')

```


```{r add cd8 majority clusters}


seur_majclust <- AddMajorityCluster(just.cd8, 'clusters')
seur_majlabs <- seur_majclust$maj.clusters; names(seur_majlabs) <- rownames(seur_majclust)
just.cd8 <- AddMetaData(just.cd8, metadata = seur_majlabs, col.name='maj.clusters')
just.cd8$maj.clusters <- ifelse(just.cd8$clonotype.family.size==1, just.cd8$clusters, just.cd8$maj.clusters)

sf_2_majclust <- AddMajorityCluster(just.cd8, 'CD8.cluster...2.clusters')
sf_2_majlabs <- sf_2_majclust$maj.CD8.cluster...2.clusters; names(sf_2_majlabs) <- rownames(sf_2_majclust)
just.cd8 <- AddMetaData(just.cd8, metadata = sf_2_majlabs, col.name='maj.sf_2_cluster')
just.cd8$maj.sf_2_cluster <- ifelse(just.cd8$clonotype.family.size==1, just.cd8$CD8.cluster...2.clusters, just.cd8$maj.sf_2_cluster)

sf_6_majclust <- AddMajorityCluster(just.cd8, 'CD8.cluster...6.clusters')
sf_6_majlabs <- sf_6_majclust$maj.CD8.cluster...6.clusters; names(sf_6_majlabs) <- rownames(sf_6_majclust)
just.cd8 <- AddMetaData(just.cd8, metadata = sf_6_majlabs, col.name='maj.sf_6_cluster')
just.cd8$maj.sf_6_cluster <- ifelse(just.cd8$clonotype.family.size==1, just.cd8$CD8.cluster...6.clusters, just.cd8$maj.sf_6_cluster)


saveRDS(just.cd8, 's_f_cd8_res045_20200909.rds')
just.cd8 = readRDS('s_f_cd8_res045_20200909.rds')
# sfs.cd8 = readRDS('s_f_cd8_res06_20200902.rds')
# sfs.cd8.old = readRDS('s_f_cd8all_res06_20200901.rds')

# colnames(sfs.cd8@meta.data)[colnames(sfs.cd8@meta.data)=='labels'] <- 'singler_labels'

# singler_majclust <- AddMajorityCluster(sfs.cd8, 'singler_labels')
# ingest_majclust <- AddMajorityCluster(sfs.cd8, 'oliveira_labels')

# majclust_labels = merge(singler_majclust, ingest_majclust, by=0)[,c('Row.names', 'maj.singler_labels', 'maj.oliveira_labels')]
# rownames(majclust_labels) <- majclust_labels$Row.names; majclust_labels$Row.names <- NULL
# 
# sfs.cd8 <- AddMetaData(sfs.cd8, metadata=majclust_labels)


```


```{r calc cd8 percent of each clone, echo=F, eval=F}
just.cd8 = readRDS('s_f_cd8_res045_20200909.rds')
# freqs = just.cd8@meta.data %>% subset(!is.na(Clonotype.family.number)) %>%# & 
                              # subset(!grepl('-NA', clonotype.family.2)) 

met = just.cd8@meta.data
met$clonotype.family.size <- ifelse(met$clonotype.family.2=='P3-45', 2, met$clonotype.family.size)


pats = unique(met$patient); pats
newpats = met %>% subset(patient==pats[1] & grepl('-NA', clonotype.family.2))
newpats$clonotype.family.3 = paste0(newpats$clonotype.family.2, '_', 1:nrow(newpats))

for (pat in pats[2:length(pats)]) {
  patsub = met %>% subset(patient==pat & grepl('-NA', clonotype.family.2))
  patsub$clonotype.family.3 = paste0(patsub$clonotype.family.2, '_', 1:nrow(patsub))
  
  newpats <- rbind(newpats, patsub)
}

View(newpats[,c('patient', 'clonotype.family.2', 'clonotype.family.3', 'clonotype.family.size')])

dim(met); dim(newpats)
met = merge(met, newpats, all=T); dim(met)

met$clonotype.family.3 <- ifelse(is.na(met$clonotype.family.3), as.character(met$clonotype.family.2), met$clonotype.family.3)

# met$clonotype.family.size <- ifelse(is.na(met$clonotype.family.size), 1, met$clonotype.family.size)

met$clonotype.family.size <- as.numeric(as.character(met$clonotype.family.size))
freqs = met %>% group_by(patient, clonotype.family.size, clonotype.family.3) %>% count()
sums = aggregate(freqs$clonotype.family.size, by=list(patient=freqs$patient), FUN=sum)
rownames(sums) <- sums$patient
sums$x <- as.numeric(as.character(sums$x))
colnames(sums) <- c('patient', 'total.number.tcrs')
show(sums)

dim(sums); dim(met)
met = merge(met, sums); dim(met)

met$patient.clone.percent <- as.numeric(as.character(met$clonotype.family.size)) / as.numeric(as.character(met$total.number.tcrs)) * 100
View(met[,c('patient', 'clonotype.family.size', 'clonotype.family.3', 'patient.clone.percent')] %>%
       group_by(patient, clonotype.family.3, clonotype.family.size, patient.clone.percent) %>% count())


# freqs = freqs %>% group_by(patient, clonotype.family.size, clonotype.family.2) %>% count()

### CONFIRM THAT THE PERCENTS ACTUALLY ADD UP TO 100
arena = met[is.na(met$patient.clone.percent),]; dim(arena)
View(arena[,c('patient', 'clonotype.family.2', 'clonotype.family.3', 'clonotype.family.size')])
sumrms <- met %>% subset(!is.na(patient.clone.percent)) %>% distinct(patient.clone.percent, clonotype.family.3, patient)
sumrms <- aggregate(sumrms$patient.clone.percent, by=list(Categorty=(sumrms$patient)), FUN=sum)
sumrms

percents.add <- met$patient.clone.percent; names(percents.add) <- rownames(met)
just.cd8 <- AddMetaData(just.cd8, metadata=percents.add, col.name='patient.clone.percent')
```

```{r make cd8 tables, echo=F, eval=F}
colnames(met) <- gsub('\\.+', '\\.', colnames(met))

tcr.info = c('CDR3.alpha.or.gamma.chain', 'CDR3.beta.or.delta.chain', 'CDR3.AA.alpha.or.gamma.chain','CDR3.AA.beta.or.delta.chain', 'alpha.gamma.V', 'alpha.gamma.J','beta.delta.V','beta.delta.J','beta.delta.D','TCRalpha.chain.ID','TCRbeta.chain.ID','clonotype.family.2','clonotype.family.size', 'patient.clone.percent','patient', 'maj.clusters', 'maj.sf_2_cluster', 'maj.sf_6_cluster')

met.tcr = met[,tcr.info]
met.tcr = met.tcr %>% subset(!is.na(patient.clone.percent))

write.table(met.tcr, 'cells_with_tcrs_info_20200909.txt', sep='\t')

met.tcr.family <- met.tcr %>% distinct(.keep_all=T)

write.table(met.tcr.family, 'tcr_families_info_20200909.txt', sep='\t')

```

```{r update metadata in the object}
rownames(met) <- met$Row.names
met.update = met[,setdiff(colnames(met), colnames(just.cd8@meta.data))]

just.cd8 <- AddMetaData(just.cd8, metadata=met.update)

saveRDS(just.cd8, 's_f_cd8_res045_20200909.rds')
just.cd8 = readRDS('s_f_cd8_res045_20200909.rds')

scanpy.labels = read.delim('just_cd8_ingested_into_tils.txt', sep='\t')
rownames(scanpy.labels) <- scanpy.labels$Row_names
scanpy.add = scanpy.labels$oliveira_labels; names(scanpy.add)<-rownames(scanpy.labels)
just.cd8 <- AddMetaData(just.cd8, metadata=scanpy.add, col.name='oliveira_labels')

oli_majclust <- AddMajorityCluster(just.cd8, 'oliveira_labels')
oli_majlabs <- oli_majclust$maj.oliveira_labels; names(oli_majlabs) <- rownames(oli_majclust)
just.cd8 <- AddMetaData(just.cd8, metadata = oli_majlabs, col.name='maj.oliveira_labels')
just.cd8$maj.oliveira_labels <- ifelse(just.cd8$clonotype.family.size==1, just.cd8$oliveira_labels, just.cd8$maj.oliveira_labels)

saveRDS(just.cd8, 's_f_cd8_res045_20200909.rds')

singler.labels = read.delim('sf_tilsscored_singler_20200910.txt', sep='\t')
singler.add = singler.labels[,c('labels', 'pruned.labels')]; colnames(singler.add) <- paste0('singler.', colnames(singler.add))
just.cd8 <- AddMetaData(just.cd8, metadata=singler.add)
fingler.majclust <- AddMajorityCluster(just.cd8, 'singler.labels')
singler_majlabs <- fingler.majclust$maj.singler.labels; names(singler_majlabs) <- rownames(fingler.majclust)
just.cd8 <- AddMetaData(just.cd8, metadata = singler_majlabs, col.name='maj.singler_labels')
just.cd8$maj.singler_labels <- ifelse(just.cd8$clonotype.family.size==1, just.cd8$singler.labels, just.cd8$maj.singler_labels)

DimPlot(just.cd8, group.by='singler.labels')
saveRDS(just.cd8, 's_f_cd8_res045_20200909.rds')

```

```{r save data for scanpy}
require(loomR)
just.cd8@meta.data[] <- lapply(just.cd8@meta.data, as.character)
just.cd8@meta.data[is.na(just.cd8@meta.data)] <- ''
just.cd8@meta.data <- just.cd8@meta.data[,c('Row.names', 'orig.ident', 'nCount_RNA', 'nFeature_RNA', 'patient', 'Sample.name', 'title')]

just.cd8.loom = as.loom(just.cd8, filename='just_cd8.loom')
just.cd8.loom$close_all()

just.cd8.meta = just.cd8@meta.data
write.table(just.cd8.meta, 's_f_cd8_res045_20200909_metadata.txt', sep='\t')

write.table(data.frame('var_genes'=just.cd8@assays$RNA@var.features), 'just_cd8_var_genes.txt', sep='\t')


require(Matrix)
writeMM(t(just.cd8@assays$RNA@counts), 'just_cd8_counts.mtx')
writeMM(just.cd8@graphs$RNA_snn, 'just_cd8_snn_neighbors.mtx')
writeMM(just.cd8@graphs$RNA_nn, 'just_cd8_nn_neighbors.mtx')

```

