---
title: "Melanoma Sample Integration"
author: "Kari Stromhaug and Giacomo Oliveira"
date: "4/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading libraries and setting path
```{r loading libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
library(ggpubr)
library(cowplot)
library(Seurat)
library(openxlsx)
library(stringr)
library(clustree)
library(harmony)
library(tidyr)
library(ggplot2)

setwd("~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/")
source('gsea_functions.R')
```


```{r read in data}
setwd("~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/")
p2.tils = readRDS('patient2/R_output/Seurat_Objects/TILs.Res04.clustered.seurat.20200228.rds')
p6.tils = readRDS('patient6/R_output/Seurat_Objects/TILs.clustered.seurat.20200306.rds')
p11.tils = readRDS('patient11/R_output/Seurat_Objects/TILs.Res03.clustered.seurat.20200309.rds')
p15.tils = readRDS('patient15/R_output/Seurat_Objects/TILs.Res05.clustered.seurat.20200302.rds')

```

```{r}
tils.all <- merge(x = p2.tils, y = c(p6.tils, p11.tils, p15.tils), add.cell.ids = c("p2", "p6", 'p11', 'p15'), project = "Melanoma TCR TILs")

tils.all.copy = tils.all
```


```{r}
origs = tils.all@meta.data[,c('orig.ident', 'sample')]; origs$s = rownames(origs)
require(tidyverse)
news = origs %>% separate (s, sep='_', into=c('patient', 'barcode'))
patient = news$patient; names(patient) = rownames(news)
tils.all = AddMetaData(tils.all, metadata=patient, col.name='patient')

unique(tils.all@meta.data[which(tils.all@meta.data$patient=='p2'), ]$dn.adt)

head(tils.all@meta.data[which(tils.all@meta.data$patient=='p6'), c('patient', 'dn.adt', 'dp.adt', 'cd4.adt', 'cd8.adt') ])


```


```{r normalization}

tils.all <- NormalizeData(tils.all, assay = "RNA", normalization.method = "LogNormalize", scale.factor = 1e6)
```

### Cluster without batch correction
```{r clustering without batch correction} 

tils.all <- tils.all %>% FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>% 
  ScaleData()

tils.all <- tils.all %>% RunPCA(features = VariableFeatures(tils.all), verbose=F)

tils.all <- tils.all %>% RunUMAP(reduction = "pca", dims = 1:50) %>% 
  FindNeighbors(dims = 1:50) %>%
  FindClusters()

DimPlot(tils.all, label=TRUE)

origs = tils.all@meta.data[,c('orig.ident', 'sample')]; origs$s = rownames(origs)
require(tidyverse)
news = origs %>% separate(s, sep='_', into=c('patient', 'barcode'))
patient = news$patient; names(patient) = rownames(news)
tils.all = AddMetaData(tils.all, metadata=patient, col.name='patient')

Idents(tils.all)=tils.all$patient
DimPlot(tils.all)

## Batch Plot
Idents(tils.all) <- tils.all@meta.data$sample
DimPlot(tils.all, reduction = "umap")

```



# Harmony

#### Look at PCs
```{r look at PCs}
tils.harmony <- tils.all %>% NormalizeData(assay = "RNA", normalization.method = "LogNormalize", scale.factor = 1e6)
tils.harmony <- tils.harmony %>% FindVariableFeatures(selection.method = "vst", nfeatures = 2000)
tils.harmony <- tils.harmony %>% ScaleData()
tils.harmony <- tils.harmony %>% RunPCA(features = VariableFeatures(tils.harmony), verbose = FALSE)

## already has the patient there
options(repr.plot.height = 5, repr.plot.width = 12)
DimPlot(object = tils.harmony, dims=c(1,2), reduction = "pca", pt.size = .1, group.by = "patient", split.by='patient')
DimPlot(object = tils.harmony, dims=c(1,2), reduction = "pca", pt.size = .1, group.by = "patient")
DimPlot(object = tils.harmony, dims=c(2,3), reduction = "pca", pt.size = .1, group.by = "patient")

VlnPlot(object = tils.harmony, features = "PC_1", group.by = "patient", pt.size = .1)
# plot_grid(p1,p3)

```


#### Run Harmony
```{r run harmony}
library(harmony)
options(repr.plot.height = 2.5, repr.plot.width = 6)
tils.harmony <- tils.harmony %>% RunHarmony("patient", plot_convergence = TRUE)

options(repr.plot.height = 5, repr.plot.width = 12)
dp1 <- DimPlot(object = tils.harmony, reduction = "harmony", pt.size = .1, group.by = "patient")
v1 <- VlnPlot(object = tils.harmony, features = "harmony_1", group.by = "patient", pt.size = .1)
plot_grid(dp1, v1)
DimPlot(object = tils.harmony, dims=c(1,2), reduction = "harmony", pt.size = .1, group.by = "patient", split.by='patient')

```


#### Cluster again
```{r recluster}

tils.harmony <- tils.harmony %>% 
    RunUMAP(reduction = "harmony", dims=1:50) %>%
    FindNeighbors(reduction = "harmony", dims = 1:50) %>% 
    FindClusters(resolution = 0.5) %>% 
    identity()

DimPlot(tils.harmony, reduction='umap', label=TRUE)
DimPlot(tils.harmony, reduction='umap', split.by='patient')

```


```{r read in tils.harmony}
saveRDS(tils.harmony, '~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/tils_combined/tils.harmonized.20200403.rds')
# tils.harmony = readRDS('~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/tils_combined/tils.harmonized.20200403.rds')
```

## Run UMAP on tils.harmony cells at different resolutions
```{r res til, message=FALSE, warning=FALSE}
res = c(0.30, 0.35, 0.40, 0.45, 0.50, 0.55, 0.60, 0.65)#, 0.70)#, 0.75, 0.80)#, 0.85, 0.90)

for (i in res) {
  tils.harmony <- FindClusters(tils.harmony, resolution=i)
  tils.harmony@meta.data[,paste0('clusters_', as.character(i))]<-tils.harmony$seurat_clusters
}

clustree(tils.harmony)

```

### Plotting different resolutions
```{r plotting tils.harmony resolutions, message=FALSE, warning=FALSE}
Idents(tils.harmony) <- tils.harmony$clusters_0.3
d1 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.30")

Idents(tils.harmony) <- tils.harmony$clusters_0.35
d2 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.35")

Idents(tils.harmony) <- tils.harmony$clusters_0.4
d3 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.40")

Idents(tils.harmony) <- tils.harmony$clusters_0.45
d4 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.45")

Idents(tils.harmony) <- tils.harmony$clusters_0.5
d5 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.50")

Idents(tils.harmony) <- tils.harmony$clusters_0.55
d6 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.55")

Idents(tils.harmony) <- tils.harmony$clusters_0.6
d7 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.60")

Idents(tils.harmony) <- tils.harmony$clusters_0.65
d8 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.65")

# Idents(tils.harmony) <- tils.harmony$clusters_0.7
# d9 = DimPlot(tils.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.70")

plot_grid(d1, d2, d3, d4)
plot_grid(d5, d6, d7, d8)

Idents(tils.harmony)<-tils.harmony$clusters_0.6
tils.harmony$clusters<-tils.harmony$clusters_0.6

```


### Look at a few features
```{r plotting various things}
p1 = FeaturePlot(tils.harmony, features='cd4.adt')
p2 = FeaturePlot(tils.harmony, features='cd8.adt')
p3 = FeaturePlot(tils.harmony, features='dp.adt')
p4 = FeaturePlot(tils.harmony, features='dn.adt')
plot_grid(p1, p2, p3, p4)

k1 = FeaturePlot(tils.harmony, features='CD19.CITEseq.norm', min.cutoff = 1)
k2 = FeaturePlot(tils.harmony, features='CD3.CITEseq.norm', min.cutoff = 1)
k3 = FeaturePlot(tils.harmony, features='CD11b.CITEseq.norm', min.cutoff = 1)
plot_grid(k1, k2, k3)
FeaturePlot(tils.harmony, feature = "FOXP3")

```


### Renaming Clonotypes
```{r}
meta = tils.harmony@meta.data
are.clones = rownames(meta[!is.na(meta$til.clonotype.family) | !is.na(meta$til.pre.clonotype.family) | !is.na(meta$til.rel.clonotype.family), ]); length(are.clones)
are.traced = rownames(meta[!is.na(meta$TCR.Clone) & meta$TCR.Clone!=0, ]); length(are.traced)

# DimPlot(tils.harmony, cells.highlight=are.traced)
# DimPlot(tils.harmony, cells.highlight=are.traced, split.by='patient')

meta$patient.tcf = paste0(meta$patient, '.', meta$til.clonotype.family)
meta$patient.tcf = ifelse(meta$patient.tcf %in% c('p6.NA', 'p2.NA', 'p11.NA', 'p15.NA'), NA, meta$patient.tcf)
meta$patient.pre.tcf = paste0(meta$patient, '.', meta$til.pre.clonotype.family)
meta$patient.pre.tcf = ifelse(meta$patient.pre.tcf %in% c('p6.NA', 'p2.NA', 'p11.NA', 'p15.NA'), NA, meta$patient.pre.tcf)
meta$patient.rel.tcf = paste0(meta$patient, '.', meta$til.rel.clonotype.family)
meta$patient.rel.tcf = ifelse(meta$patient.rel.tcf %in% c('p6.NA', 'p2.NA', 'p11.NA', 'p15.NA'), NA, meta$patient.rel.tcf)

patient.tcf.freqs = meta %>% subset(!is.na(patient.tcf)) %>% group_by(patient.tcf) %>% tally(name='tcf.counts'); dim(patient.tcf.freqs)
patient.pre.tcf.freqs = meta %>% subset(!is.na(patient.pre.tcf)) %>% group_by(patient.pre.tcf) %>% tally(name='tcf.pre.counts'); dim(patient.pre.tcf.freqs)
patient.rel.tcf.freqs = meta %>% subset(!is.na(patient.rel.tcf)) %>% group_by(patient.rel.tcf) %>% tally(name='tcf.rel.counts'); dim(patient.rel.tcf.freqs)

dim(meta); head(rownames(meta))
meta$cell.barcode = rownames(meta)
# meta = m
meta = merge(meta, patient.tcf.freqs, by=c('patient.tcf'), all.x=TRUE); dim(meta)
meta = merge(meta, patient.pre.tcf.freqs, by=c('patient.pre.tcf'), all.x=TRUE); dim(meta)
meta = merge(meta, patient.rel.tcf.freqs, by=c('patient.rel.tcf'), all.x=TRUE); dim(meta)
meta[is.na(meta)]<-0


##### create a new clonotype ID so that I don't have 3 columns
meta$final.clonotype.family <- meta$patient.tcf
meta$p11.fams = paste0(meta$patient.pre.tcf, '-', meta$patient.rel.tcf)
meta$final.clonotype.family <- ifelse(meta$p11.fams!='0-0', meta$p11.fams, meta$final.clonotype.family)
final.freqs = meta %>% subset(final.clonotype.family!=0) %>% group_by(final.clonotype.family) %>% tally(name='final.family.counts'); dim(final.freqs)
meta = merge(meta, final.freqs, by='final.clonotype.family', all.x=TRUE); dim(meta)

rownames(meta)<-meta$cell.barcode
final.family = as.character(meta$final.clonotype.family); names(final.family)<-rownames(meta)
final.family.counts = as.character(meta$final.family.counts); names(final.family.counts)<-rownames(meta)
tils.harmony <- AddMetaData(tils.harmony, metadata=final.family, col.name='final.clonotype.family')
tils.harmony <- AddMetaData(tils.harmony, metadata=final.family.counts, col.name='final.family.counts')
       
```


#Save and reload the data
```{r save data}
saveRDS(tils.harmony, 'Seurat_Objects/tils.harmonized.20200403.rds')
tils.harmony = readRDS('Seurat_Objects/tils.harmonized.20200403.rds')
```

# Find markers for each cluster
```{r top til.cd8 markers}
# find markers for every cluster compared to all remaining cells, report only the positive ones
til.markers <- FindAllMarkers(tils.harmony, min.pct = 0.25, logfc.threshold = log(2))
top_10 = til.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)

top_10 = data.frame(top_10)
show(top_10)
write.table(top_10, 'harmonizedtils_top10genes_by_cluster.xlsx', sep=',')

png("./Plots/harmonizedtils_top10_heatmap.png", width = 10, height = 7, units='in', res=500)
DoHeatmap(tils.harmony, features = top_10$gene) + font("xy.text", size = 4)
dev.off()


top20 <- til.markers %>% group_by(cluster) %>% top_n(n = 20, wt = avg_logFC)
png("./plots/harmonizedtils_top20_heatmap.png", width = 10, height = 8, units='in', res=500)
DoHeatmap(tils.harmony, features = top20$gene) + font("xy.text", size = 3) 
dev.off()

write.table(til.markers, file='harmonizedtils.markers.xlsx', sep=',')
```


#Get table clonotype distribution within clusters 
```{r echo=FALSE, eval=FALSE}
g.table = tils.harmony@meta.data[,c('orig.ident', 'sample', 'TCR.Clone', 'til.clonotype.family', 'final.clonotype.family', 'final.family.counts', 'til.pre.clonotype.family', 'til.rel.clonotype.family', 'final.clonotype.family', 'patient', 'clusters','seurat_clusters')]
g.table$cell.barcode <- rownames(tils.harmony@meta.data)
rownames(g.table) <- rownames(tils.harmony@meta.data)


wb <- createWorkbook()
name <- "TCR clones"
addWorksheet(wb, name)
writeData(wb, sheet = name, g.table)

saveWorkbook(wb, file = 'tils.harmony.clonotype.cluster.spreadsheet.xlsx', overwrite = TRUE)

```





# Just CD8 TILS
```{r subset out tils}

cells <- rownames(tils.harmony@meta.data)[(tils.harmony@meta.data$cd8.adt==1 & tils.harmony@meta.data$cd4.adt==0)]; length(cells)
tils.CD8.harmony <- subset(tils.harmony, cells = cells)
tils.CD8.harmony <- NormalizeData(tils.CD8.harmony, assay = "RNA", normalization.method = "LogNormalize", scale.factor = 1e6)

```
### Check selection
```{r plotting various things}
p1 = FeaturePlot(tils.CD8.harmony, features='cd4.adt')
p2 = FeaturePlot(tils.CD8.harmony, features='cd8.adt')
p3 = FeaturePlot(tils.CD8.harmony, features='dp.adt')
p4 = FeaturePlot(tils.CD8.harmony, features='dn.adt')
plot_grid(p1, p2, p3, p4)
DimPlot(tils.harmony, reduction='umap', label=TRUE)
DimPlot(tils.CD8.harmony, reduction='umap', label=TRUE)


```


#### Cluster again
```{r recluster}

tils.CD8.harmony <- tils.CD8.harmony %>% 
    RunUMAP(reduction = "harmony", dims=1:50) %>%
    FindNeighbors(reduction = "harmony", dims = 1:50) %>% 
    FindClusters(resolution = 0.6) %>% 
    identity()

DimPlot(tils.CD8.harmony, reduction='umap', label=TRUE)
DimPlot(tils.CD8.harmony, reduction='umap', split.by='patient')
```

#Save and reload the data
```{r save data}
saveRDS(tils.CD8.harmony, 'Seurat_Objects/tils.CD8.harmonized.20200407.rds')
tils.CD8.harmony = readRDS('Seurat_Objects/tils.CD8.harmonized.20200407.rds')
saveRDS(tils.CD8.harmony, 'Seurat_Objects/tils.CD8.R0.6.harmonized.20200422.rds')
tils.CD8.harmony = readRDS('Seurat_Objects/tils.CD8.R0.6.harmonized.20200422.rds')
tils.CD8.harmony$clusters <- tils.CD8.harmony$seurat_clusters
```


## Run UMAP on tils.CD8.harmony cells at different resolutions
```{r res til, message=FALSE, warning=FALSE}
res = c(0.30, 0.35, 0.40, 0.45, 0.50, 0.55, 0.60, 0.65, 0.70, 0.75, 0.80, 0.85, 0.90, 0.95, 1)

for (i in res) {
  tils.CD8.harmony <- FindClusters(tils.CD8.harmony, resolution=i)
  tils.CD8.harmony@meta.data[,paste0('clusters_', as.character(i))]<-tils.CD8.harmony$seurat_clusters
}

clustree(tils.CD8.harmony)

```

### Plotting different resolutions
```{r plotting tils.harmony resolutions, message=FALSE, warning=FALSE}
Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.3
d1 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.30")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.35
d2 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.35")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.4
d3 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.40")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.45
d4 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.45")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.5
d5 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.50")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.55
d6 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.55")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.6
d7 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.60")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.65
d8 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.65")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.7
d9 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.70")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.75
d10 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.75")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.8
d11 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.80")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.85
d12 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.85")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.9
d13 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.90")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_0.95
d14 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.95")

Idents(tils.CD8.harmony) <- tils.CD8.harmony$clusters_1
d15 = DimPlot(tils.CD8.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 1")


plot_grid(d1, d2)
plot_grid(d3, d4)
plot_grid(d5, d6)
plot_grid(d7, d8)
plot_grid(d9, d10)
plot_grid(d11,d12)
plot_grid(d15,d14, d15)

```

#### Finding markers and making heatmaps
```{r find markers genes for each cluster, paged.print=TRUE,fig.height=3, fig.width=8}
tils = readRDS('~/Dropbox (Partners HealthCare)/Melanoma P2 Analysis/tils_combined/Seurat_Objects/tils.CD8.R0.6.harmonized.20200422.rds')

markers <- FindAllMarkers(tils)


## find the top genes for each cluster
top.genes <- markers %>% group_by(cluster) %>% top_n(50, wt=avg_logFC)

## changed out CD45RA and CD45RO with PTPRC
genes <- c('CD4','CD3E','CD8A',
           'SELL','CCR7','IL7R','CD28','PTPRC','FAS','CD27','ITGAE',
           'PDCD1','TIGIT','HAVCR2','LAG3','CTLA4','VTCN1','CD244','KLRG1','TNFRSF14','BTLA',
           'ENTPD1','CD69','IL2RA','ICOS','TNFRSF4','TNFRSF9','HLA-DRA',
           'GZMA','GZMB','GZMH','GZMK','GZMM','PRF1','IFNG','FASLG','TNF',
           'TCF7','EOMES','TBX21','PRDM1','TOX','ZNF683','FOXP3','NR4A1','MKI67',
           'TRGV9','TRDV2','KLRB1','KLRC3')

exp <- tils@assays$RNA@data[genes,]
sc <- tils@assays$RNA@scale.data[intersect(genes, rownames(tils@assays$RNA@scale.data)),]
tils@meta.data$cell.barcode <- rownames(tils@meta.data)
exp.t <- t(as.matrix(exp))
exp.t <- cbind(tils@meta.data[,c('cell.barcode','seurat_clusters','patient')], exp.t)
exp.tm <- melt(exp.t, value.name='expression')

sc.t <- t(sc)
sc.t <- cbind(tils@meta.data[,c('cell.barcode','seurat_clusters','patient')], sc.t)
sc.tm <- melt(sc.t, value.name='expression')

exp.cl0 <- exp.tm %>% subset(seurat_clusters==0); dim(exp.cl0)
ggplot(exp.cl0) + geom_boxplot(aes(x=variable, y=expression)) + theme(axis.text.x = element_text(angle = 90))

ggplot(exp.tm) + geom_violin(aes(x=variable, y=expression)) + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows='seurat_clusters')

### becaue of the outliers, removing the 5% highest and 5% lowest cells for each category
exp.f <- data.frame()
for (i in 0:13) {
  sub <- sc.tm %>% subset(seurat_clusters == i)
  sub <- sub[order(sub$expression), ]
  fivep <- nrow(sub) * 0.05
  print(dim(sub))
  sub <- sub[fivep:(nrow(sub)-fivep), ]
  print(dim(sub))
  exp.f <- rbind(exp.f, sub)
  cat('\n')
}

exp.sd <- data.frame()
for (i in 0:13) {
  sub <- sc.tm %>% subset(seurat_clusters == i)
  sub <- sub[order(sub$expression), ]
  mn <- mean(sub$expression)
  std <- 2*sd(sub$expression)
  print(dim(sub))
  sub <- sub %>% subset(expression < (mn+std) & expression > (mn-std))
  print(dim(sub))
  exp.sd <- rbind(exp.sd, sub)
  cat('\n')
}

ggplot(exp.f) + geom_violin(aes(x=variable, y=expression)) + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows='seurat_clusters')

ggplot(exp.sd) + geom_violin(aes(x=variable, y=expression)) + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows='seurat_clusters')

ss <- exp.sd %>% subset(seurat_clusters %in% c(0,1,3))
ggplot(ss) + geom_violin(aes(x=variable, y=expression)) + theme(axis.text.x = element_text(angle = 90)) + facet_grid(rows='seurat_clusters')

require(ggplot2)

```




#### DGE testing in the tils.CD8.harmony population 
```{r de testing tils.CD8.harmony, paged.print=TRUE,fig.height=3, fig.width=8}
source('gsea_functions.R')
all_genes <- de_test(tils.CD8.harmony, clusters = levels(tils.CD8.harmony$seurat_clusters))
de_genes <- all_genes[all_genes$p_val_adj < 0.05]
de_genes

FindMarkers(tils.CD8.harmony, ident.1 = 1, ident.2 = 2, verbose = FALSE, logfc.threshold = 0, min.pct = 0)

# Making plots for a few of the top genes
plots <- de_violin(seurat = tils.CD8.harmony, de_genes = de_genes, row = c(1,2,3,9))
CombinePlots(plots = plots, ncol = 4)
```


#### Perform GSEA on DE genes that are down-/upregulated in populations in each cluster
```{r GSEA tils.CD8.harmony, warning=FALSE, paged.print=TRUE}


m_df = msigdbr(species = "Homo sapiens", category = "H")
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)

apoptosis = read.delim('~/Downloads/geneset.txt', sep='\t')
apoptosis_list = list('apoptosis'=apoptosis[2:nrow(apoptosis), 1])
# GSEA for genes upregulated in cGVHD - within clusters
gsea_res_tils.CD8.harmony <- gsea(all_genes, levels(tils.CD8.harmony$seurat_clusters), 'tils.CD8.harmony', apoptosis_list)

# Printing some of the top gene sets (based on p.adj.)
gsea_res_tils.CD8.harmony[gsea_res_tils.CD8.harmony$padj<0.05,]
```

GSEA tests were performed using "fgsea" (https://bioconductor.org/packages/devel/bioc/manuals/fgsea/man/fgsea.pdf). 
Quick explanation of some columns in the output:

1. NES – enrichment score normalized to mean enrichment of random samples of the same size (when negative; all genes in "leadingEdge" are *down*regulated in cGVHD compared to BM - and vice versa).

2. nMoreExtreme – a number of times a random gene set had a more extreme enrichment score value.

3. size – size of the pathway after removing genes not present in ‘names(stats)‘.

4. leadingEdge – vector with indexes of leading edge genes that drive the enrichment. Can only be seen in Excel documents available in the Dropbox.



# Find markers for each cluster
```{r top til.cd8 markers}
# find markers for every cluster compared to all remaining cells, report only the positive ones
til.CD8.markers <- FindAllMarkers(tils.CD8.harmony, min.pct = 0.25, logfc.threshold = log(2))
top_10 = til.CD8.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)

top_10 = data.frame(top_10)
show(top_10)
write.table(top_10, 'harmonizedtilsCD8R06_top10genes_by_cluster.xlsx', sep=',')

png("./Plots/harmonizedtilsCD8R06_top10_heatmap.png", width = 10, height = 7, units='in', res=500)
DoHeatmap(tils.CD8.harmony, features = top_10$gene) + font("xy.text", size = 4)
dev.off()


top20 <- til.CD8.markers %>% group_by(cluster) %>% top_n(n = 20, wt = avg_logFC)
png("./Plots/harmonizedtilsCD8R06_top20_heatmap.png", width = 10, height = 8, units='in', res=500)
DoHeatmap(tils.CD8.harmony, features = top20$gene) + font("xy.text", size = 3) 
dev.off()

write.table(til.CD8.markers, file='harmonizedtilsCD8R06.markers.xlsx', sep=',')
```

# Make Heatmaps with selected genes
```{r top til.cd8 markers}
# find markers for every cluster compared to all remaining cells, report only the positive ones
genes = c('CD4', 'CD8A','GZMK', 'GZMM', 'PRF1', 'IFNG', 'FASLG', 
             'TNF', 'CD69', 'ICOS', 'TNFRSF9', 'HLA-DRA', 'ENTPD1')

Idents(tils.CD8.harmony)<-tils.CD8.harmony$clusters
png(" ## FILE ## ", width = 10, height = 7, units='in', res=500)
DoHeatmap(tils.CD8.harmony, features = genes, slot='data', lines.width = 300) + #font("xy.text", size = 4) + 
  #theme(legend.position = "none") +
  scale_fill_gradientn(colors = c("blue", rep('chartreuse4', 20), rep('chartreuse3', 3), rep('chartreuse2', 3), rep('chartreuse1', 3), "yellow"))
dev.off()


```

#Get table clonotype distribution within clusters 
```{r echo=FALSE, eval=FALSE}
g.table = tils.CD8.harmony@meta.data[,c('orig.ident', 'sample', 'TCR.Clone', 'til.clonotype.family', 'final.clonotype.family', 'final.family.counts', 'til.pre.clonotype.family', 'til.rel.clonotype.family', 'patient', 'clusters','seurat_clusters')]
g.table$cell.barcode <- rownames(tils.CD8.harmony@meta.data)
rownames(g.table) <- rownames(tils.CD8.harmony@meta.data)


wb <- createWorkbook()
name <- "TCR clones"
addWorksheet(wb, name)
writeData(wb, sheet = name, g.table)

saveWorkbook(wb, file = 'tilsCD8.R06.harmony.clonotype.cluster.spreadsheet.xlsx', overwrite = TRUE)

```

score clusters
```{r echo=FALSE, eval=FALSE}
tils.CD8.harmony = readRDS('Seurat_Objects/tils.CD8.R0.6.harmonized.20200422.rds')

t_eff = list(c('GNLY','NKG7','GZMH','CST7','KLRD1','GZMB','CCL5','PRF1','IFITM2','IFITM3','LITAF','HCST','ITGB2','GZMA','GPR56','GZMM','IFITM1','CD52','RAP1B','ZNF683'))
t_em = list(c('GZMK','CCL4','CCL5','CST7','DUSP2','CD74','GZMA','NKG7','CCL3','CD8B','IFNG','CD8A','LOC100996809','HLA-A','CXCR4','GZMM','CLDND1','HLA-DRB4','DNAJB1','GZMH'))
mitosis = list(c('HIST1H4C','TUBB','TUBA1B','HMGN2','GAPDH','H2AFZ','HMGB1','HMGB2','TPI1','PFN1','ACTB','PPIA','H2AFV','DUT','CALM2','GSTP1','HMGN1','SLC25A5','CRIP1','ENO1'))

Idents(tils.CD8.harmony)<-tils.CD8.harmony$seurat_clusters
cell_group = mitosis
cell_group[[1]] = cell_group[[1]][cell_group[[1]] %in% rownames(tils.CD8.harmony)]
tils.CD8.harmony <- AddModuleScore(tils.CD8.harmony, features=cell_group, name='mitosis_')
unique(tils.CD8.harmony$seurat_clusters)
unique(tils.CD8.harmony$clusters)
ids = grep('Cluster[0-9]+', colnames(tils.CD8.harmony@meta.data))
tils.CD8.harmony@meta.data[,ids]<-NULL
cls = tils.CD8.harmony@meta.data[,ids]

DimPlot(tils.CD8.harmony, label=TRUE)
scores = tils.CD8.harmony@meta.data[,c('seurat_clusters', 'Group_1', 't_em_1', 'mitosis_1')]
cluster_scores = aggregate(scores, list(scores$seurat_clusters), mean) 
## T eff: clusters 0, 10, 11, 7 all have scores above 1
## T em: clusters 13 and 8 have scores above 2
## mitosis: cluster 5 has a high score (2.99)

```



```{r save matrix file}
tils.CD8.harmony = readRDS('Seurat_Objects/tils.CD8.R0.6.harmonized.20200422.rds')
DimPlot(tils.CD8.harmony)

library(Matrix)
writeMM(obj = t(tils.CD8.harmony@assays$RNA@counts), file="cd8_counts_sparse_matrix_t.mtx")
meta = tils.CD8.harmony@meta.data[,c('seurat_clusters', 'final.family.counts', 'final.clonotype.family', 'patient', 'TCR.Clone', 'percent_mito', 'nCount_RNA')]
write.table(meta, 'tils_cd8_meta.txt', sep='\t', row.names=F, quote=F)
genes = data.frame('genes'=rownames(tils.CD8.harmony))
write.table(genes, 'tils_cd8_genes.txt', sep='\t', quote=F, row.names=F)
cells = data.frame('cells'=rownames(meta))
write.table(cells, 'tils_cd8_cells.txt', sep='\t', row.names = F, quote=F)

## get just cluster 4
cd8.12 = subset(tils.CD8.harmony, idents='12')
FeaturePlot(cd8.12, features=c('CD45RA.CITEseq.norm','CD62L.CITEseq.norm'), min.cutoff = 1.5)
m = cd8.12@meta.data
high = m %>% subset(CD45RA.CITEseq.norm > 1.5 & CD62L.CITEseq.norm > 1.5); dim(high)
cd8.12.high = subset(cd8.12, cells=rownames(high))
FeaturePlot(cd8.12.high, features=c('CD45RA.CITEseq.norm','CD62L.CITEseq.norm'), min.cutoff = 1.5)
rownames(high)

```



# Just CD4 TILS
```{r subset out tils}

cells <- rownames(tils.harmony@meta.data)[(tils.harmony@meta.data$cd4.adt==1 & tils.harmony@meta.data$cd8.adt==0)]; length(cells)
tils.CD4.harmony <- subset(tils.harmony, cells = cells)
tils.CD4.harmony <- NormalizeData(tils.CD4.harmony, assay = "RNA", normalization.method = "LogNormalize", scale.factor = 1e6)

```
### Check selection
```{r plotting various things}
p1 = FeaturePlot(tils.CD4.harmony, features='cd4.adt')
p2 = FeaturePlot(tils.CD4.harmony, features='cd8.adt')
p3 = FeaturePlot(tils.CD4.harmony, features='dp.adt')
p4 = FeaturePlot(tils.CD4.harmony, features='dn.adt')
plot_grid(p1, p2, p3, p4)
DimPlot(tils.harmony, reduction='umap', label=TRUE)
DimPlot(tils.CD4.harmony, reduction='umap', label=TRUE)


```


#### Cluster again
```{r recluster}

tils.CD4.harmony <- tils.CD4.harmony %>% 
    RunUMAP(reduction = "harmony", dims=1:50) %>%
    FindNeighbors(reduction = "harmony", dims = 1:50) %>% 
    FindClusters(resolution = 0.6) %>% 
    identity()

DimPlot(tils.CD4.harmony, reduction='umap', label=TRUE)
DimPlot(tils.CD4.harmony, reduction='umap', split.by='patient')
```

#Save and reload the data
```{r save data}
saveRDS(tils.CD4.harmony, 'Seurat_Objects/tils.CD4.harmonized.20200514.rds')
tils.CD4.harmony = readRDS('Seurat_Objects/tils.CD4.harmonized.20200514.rds')
tils.CD4.harmony$clusters <- tils.CD4.harmony$seurat_clusters
library(Matrix)
writeMM(obj = t(tils.CD4.harmony@assays$RNA@counts), file="cd4_counts_sparse_matrix_t.mtx")
meta = tils.CD4.harmony@meta.data[,c('seurat_clusters', 'final.family.counts', 'final.clonotype.family', 'patient', 'TCR.Clone')]
write.table(meta, 'tils_cd4_meta.txt', sep='\t', row.names=F, quote=F)
genes = data.frame('genes'=rownames(tils.CD4.harmony))
write.table(genes, 'tils_cd4_genes.txt', sep='\t', quote=F, row.names=F)

## get just cluster 4
cd4.4 = subset(tils.CD4.harmony, idents='4')
FeaturePlot(cd4.4, features=c('CD45RA.CITEseq.norm','CD62L.CITEseq.norm'), min.cutoff = 1.5)
m = cd4.4@meta.data
high = m %>% subset(CD45RA.CITEseq.norm > 1.5 & CD62L.CITEseq.norm > 1.5); dim(high)
cd4.4.high = subset(cd4.4, cells=rownames(high))
FeaturePlot(cd4.4.high, features=c('CD45RA.CITEseq.norm','CD62L.CITEseq.norm'), min.cutoff = 1.5)
rownames(high)
```



## Run UMAP on tils.CD4..harmony cells at different resolutions
```{r res til, message=FALSE, warning=FALSE}
res = c(0.30, 0.35, 0.40, 0.45, 0.50, 0.55, 0.60, 0.65, 0.70, 0.75, 0.80, 0.85, 0.90, 0.95, 1)

for (i in res) {
  tils.CD4.harmony <- FindClusters(tils.CD4.harmony, resolution=i)
  tils.CD4.harmony@meta.data[,paste0('clusters_', as.character(i))]<-tils.CD4.harmony$seurat_clusters
}

clustree(tils.CD4.harmony)

```

### Plotting different resolutions
```{r plotting tils.harmony resolutions, message=FALSE, warning=FALSE}
Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.3
d1 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.30")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.35
d2 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.35")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.4
d3 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.40")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.45
d4 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.45")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.5
d5 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.50")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.55
d6 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.55")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.6
d7 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.60")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.65
d8 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.65")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.7
d9 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.70")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.75
d10 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.75")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.8
d11 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.80")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.85
d12 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.85")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.9
d13 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.90")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_0.95
d14 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 0.95")

Idents(tils.CD4.harmony) <- tils.CD4.harmony$clusters_1
d15 = DimPlot(tils.CD4.harmony, reduction = "umap", label = TRUE) + labs(title = "Res = 1")


plot_grid(d1, d2)
plot_grid(d3, d4)
plot_grid(d5, d6)
plot_grid(d7, d8)
plot_grid(d9, d10)
plot_grid(d11,d12)
plot_grid(d15,d14, d15)

```



# Find markers for each cluster
```{r top til.cd4 markers}
# find markers for every cluster compared to all remaining cells, report only the positive ones
til.CD4.markers <- FindAllMarkers(tils.CD4.harmony, min.pct = 0.25, logfc.threshold = log(2))
top_10 = til.CD4.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)

top_10 = data.frame(top_10)
show(top_10)
write.table(top_10, 'harmonizedtilsCD4_top10genes_by_cluster.xlsx', sep=',')

png("./Plots/harmonizedtilsCD4_top10_heatmap.png", width = 10, height = 7, units='in', res=500)
DoHeatmap(tils.CD4.harmony, features = top_10$gene) + font("xy.text", size = 4)
dev.off()


top20 <- til.CD4.markers %>% group_by(cluster) %>% top_n(n = 20, wt = avg_logFC)
png("./Plots/harmonizedtilsCD4_top20_heatmap.png", width = 10, height = 8, units='in', res=500)
DoHeatmap(tils.CD4.harmony, features = top20$gene) + font("xy.text", size = 3) 
dev.off()

write.table(til.CD4.markers, file='harmonizedtilsCD4.markers.xlsx', sep=',')
```



# Make Heatmaps with selected genes
```{r top til.cd4 markers}
# find markers for every cluster compared to all remaining cells, report only the positive ones
genes = c('CD4', 'CD8A','GZMK', 'GZMM', 'PRF1', 'IFNG', 'FASLG', 
             'TNF', 'CD69', 'ICOS', 'TNFRSF9', 'HLA-DRA', 'ENTPD1')

Idents(tils.CD4.harmony)<-tils.CD8.harmony$clusters
png(" ## FILE ## ", width = 10, height = 7, units='in', res=500)
DoHeatmap(tils.CD4.harmony, features = genes, slot='data', lines.width = 300) + #font("xy.text", size = 4) + 
  #theme(legend.position = "none") +
  scale_fill_gradientn(colors = c("blue", rep('chartreuse4', 20), rep('chartreuse3', 3), rep('chartreuse2', 3), rep('chartreuse1', 3), "yellow"))
dev.off()


```


#Get table clonotype distribution within clusters 
```{r echo=FALSE, eval=FALSE}
g.table = tils.CD4.harmony@meta.data[,c('orig.ident', 'sample', 'TCR.Clone', 'til.clonotype.family', 'final.clonotype.family', 'final.family.counts', 'til.pre.clonotype.family', 'til.rel.clonotype.family', 'patient', 'clusters','seurat_clusters')]
g.table$cell.barcode <- rownames(tils.CD4.harmony@meta.data)
rownames(g.table) <- rownames(tils.CD4.harmony@meta.data)


wb <- createWorkbook()
name <- "TCR clones"
addWorksheet(wb, name)
writeData(wb, sheet = name, g.table)

saveWorkbook(wb, file = 'tilsCD4.harmony.clonotype.cluster.spreadsheet.xlsx', overwrite = TRUE)

```





